---
title: Get Started
vignette: >
  %\VignetteIndexEntry{Get Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
author: "`r 'Author: Matthew Hung'`"
date: "`r paste('Last Updated:', format(Sys.Date(), '%Y/%m/%d'))`"
output:
  html_document:
    code_folding: hide
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| vscode: {languageId: r}
# load deseq2pip  
suppressPackageStartupMessages(library(deseq2pip))
```

## Run deseq2pip on RNA-seq data

Create a directory for results, currently set to tempdir()
```{r}
#| vscode: {languageId: r}
save_dir <- "/Users/hungm/The Francis Crick Dropbox/CaladoD/Matthew/git-repos/r-pipelines/deseq2pip/tests/rnaseq/pipeline/"
#save_dir <- tempdir()
dir.create(save_dir, recursive = TRUE)
```

Load DESeq2 object processed by nf-core/rnaseq.
```{r}
#| vscode: {languageId: r}
rdata <- system.file("data", "GSE189410.dds.RData", package = "deseq2pip")
tx2gene <- gzfile(system.file("data", "GSE189410.tx2gene.tsv.gz", package = "deseq2pip"))
dds <- import_nfcore_rna(rdata = rdata, tx2gene = tx2gene)
```

Setup dds comparisons and design.
```{r}
#| vscode: {languageId: r}
# set correct group factors
dds$Group2 <- factor(dds$Group2, c('IgM', 'IgG', 'IgA'))
dds$Group3 <- factor(dds$Group3, c(1:4))

# set design
design(dds) <- ~ Group3 + Group2
```

Run RNA pipeline.
```{r}
#| vscode: {languageId: r}
dds <- run_rna_pip(
    dds,
    org = "mouse",
    var = "Group2",
    design = "~ Group3 + Group2",
    batch = "Group3",
    remove_xy = TRUE,
    remove_mt = TRUE,
    min_count = 10,
    pals = NULL,
    order = "pxfc",
    save_dir = save_dir)

dds # return dds object
```

## Run deseq2pip on ATAC-seq data

Create a directory for results, currently set to tempdir()
```{r}
#| vscode: {languageId: r}
#save_dir <- "/Users/hungm/The Francis Crick Dropbox/CaladoD/Matthew/git-repos/r-pipelines/deseq2pip/tests/atacseq/pipeline/"
save_dir <- tempdir()
dir.create(save_dir, recursive = TRUE)
```

Load DESeq2 object processed by nf-core/atacseq
```{r}
rdata <- system.file("data", "GSE224512.dds.RData", package = "deseq2pip")
annotatePeaks <- gzfile(system.file("data", "GSE224512.annotatePeaks.txt.gz", package = "deseq2pip"))
dds <- import_nfcore_atac(rdata = rdata, annotatePeaks = annotatePeaks)
```

Setup dds comparisons and design.
```{r}
# subset groups for brevity
dds <- dds[, dds$Group1 %in% c("WT", 'BC', 'BCK')]

# set correct group factors
dds$Group1 <- factor(dds$Group1, c("WT", 'BC', 'BCK'))
dds$BC <- ifelse(dds$Group1 == "WT", 0, 1)
dds$BC <- factor(dds$BC, c(0, 1))
dds$K <- ifelse(dds$Group1 == "BCK", 1, 0)
dds$K <- factor(dds$K, c(0, 1))

# set design
design(dds) <- ~ BC + K
```


Run ATAC pipeline.
```{r}
#| vscode: {languageId: r}
dds <- run_atac_pip(
    dds,
    org = "mouse",
    var = "BC",
    design = "~ BC + K",
    batch = NULL,
    comparisons = NULL,
    remove_xy = TRUE,
    remove_mt = TRUE,
    min_count = 10,
    order = "pxfc",
    TSS = TRUE,
    save_dir = save_dir)

dds # return dds object
```