---
title: RNA-seq Analysis Guide
vignette: >
  %\VignetteIndexEntry{RNA-seq Analysis Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
author: "`r 'Author: Matthew Hung'`"
date: "`r paste('Last Updated:', format(Sys.Date(), '%Y/%m/%d'))`"
output:
  html_document:
    code_folding: hide
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

First, load the deseq2pip package:

```{r}
#| vscode: {languageId: r}
suppressPackageStartupMessages(library(deseq2pip))
```

## Data Import and Setup

Let's import and prepare your data. We'll use the example dataset GSE189410 to demonstrate the workflow:
```{r}
#| vscode: {languageId: r}
# load example DESeq2 object
rdata <- system.file("data", "GSE189410.dds.RData", package = "deseq2pip")
tx2gene <- gzfile(system.file("data", "GSE189410.tx2gene.tsv.gz", package = "deseq2pip"))
dds <- import_nfcore_rna(rdata = rdata, tx2gene = tx2gene)


# set group variable (optional to manually set reference level)
dds$Group2 <- factor(dds$Group2, c('IgM', 'IgG', 'IgA'))
dds$Group3 <- factor(dds$Group3, c(1:4))

# set dds design with desired variables
# for more information about setting design matrix in DESeq2, see: https://github.com/tavareshugo/tutorial_DESeq2_contrasts?tab=readme-ov-file
design(dds) <- ~ Group3 + Group2
```


## Detailed Overview of DESeq2 pipeline

The run_deseq2_pip() function is a wrapper function that runs the complete RNA-seq analysis for **every PAIRWISE and ONE-TO-ALL comparison pairs** in the specified group variable.

The analysis for each comparison pair can be broken down into three modules:

1. Quality Control: Filters genes, generates QC/PCA/distance plots, and saves expression data
* `run_qc_pip()`
* `run_dist_pip()`
2. Differential Expression Analysis: Performs DESeq2 analysis, functional gene annotation, and creates volcano plots
* `run_diffexp_pip()`
* `plot_peak_annot_pip()` (ATAC-seq only)
3. GSEA Analysis: Find enriched pathways using MSigDB collections and prepare results for EnrichmentMap visualization
* `run_gsea_pip()`
* `run_enrichmentmap_pip()`

For our example dataset, we have 3 comparison pairs in the group variable `Group2`: `IgM` vs `IgG`, `IgM` vs `IgA`, and `IgG` vs `IgA`.

Here's how to run the complete pipeline with a single command:
*Note: Repeating run_deseq2_pip will overwrite the existing files if the same directory name is used. To prevent this, please specify a different save_dir directory.*

First, set the global arguments.
```{r}
#| vscode: {languageId: r}
# set pipeline arguments
dds <- dds # required
var <- "Group2" # required
org <- "mouse" # required
design <- "~ Group3 + Group2" # required
comparisons <- c("IgG_vs_IgM", "IgA_vs_IgM", "IgA_vs_IgG", "IgM_vs_IgA+IgG", "IgG_vs_IgA+IgM", "IgA_vs_IgG+IgM") # optional
batch <- "Group3" # optional
remove_xy <- FALSE # optional
remove_mt <- FALSE # optional
min_count <- 10 # optional
pals <- NULL # optional
order <- "pxfc" # optional
save_dir <- "/Users/hungm/The Francis Crick Dropbox/CaladoD/Matthew/git-repos/r-pipelines/deseq2pip/tests/rnaseq/pipeline/"
dir.create(save_dir, recursive = TRUE)
```

```{r}
#| vscode: {languageId: r}
# run deseq2 pipeline
run_rna_pip(dds = dds, org = org, var = var, design = design, batch = batch, comparisons = NULL, remove_xy = remove_xy, remove_mt = remove_mt, min_count = min_count, pals = pals, order = order, save_dir = save_dir)
```

## Run individual modules

After an initial run of the DESeq2 pipeline, you may wish to run a specific module using different parameters or comparison groups without repeating the entire pipeline.

Here's how to run each module separately. 

Below are the commands to run each individual module:

```{r}
#| vscode: {languageId: r}
save_dir <- "/Users/hungm/The Francis Crick Dropbox/CaladoD/Matthew/git-repos/r-pipelines/deseq2pip/tests/rnaseq/pipeline/"
dir.create(save_dir, recursive = TRUE)
```
```{r}
#| vscode: {languageId: r}
# Module 1: Quality Control
dds <- run_qc_pip(dds = dds, org = org, var = var, remove_xy = remove_xy, remove_mt = remove_mt, min_count = min_count, save_dir = save_dir)
dds <- run_dist_pip(dds = dds, var = var, batch = batch, pals = pals, save_dir = save_dir)
```

```{r}
#| vscode: {languageId: r}
# Module 2: Differential Expression Analysis
res <- run_diffexp_pip(dds, org = org, var = var, design = design, comparisons = NULL, order = order, save_dir = paste0(save_dir, "/", var))
```

```{r}
#| vscode: {languageId: r}
# Module 3: GSEA Analysis
gsea <- run_gsea_pip(res = res, org = org, save_dir = paste0(save_dir, "/", var))
```

```{r}
#| vscode: {languageId: r}
# Module 4: EnrichmentMap
enrichmentmap_pip(dds, org = org, var = var, group_dir = paste0(save_dir, "/", var), save_dir = paste0(save_dir, "/", var))
```