---
title: RNA-seq Analysis Guide
vignette: >
  %\VignetteIndexEntry{RNA-seq Analysis Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
author: "`r 'Author: Matthew Hung'`"
date: "`r paste('Last Updated:', format(Sys.Date(), '%Y/%m/%d'))`"
output:
  html_document:
    code_folding: hide
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

First, load the deseq2pip package:

```{r}
#| vscode: {languageId: r}
suppressPackageStartupMessages(library(deseq2pip))
save_dir <- tempdir()
dir.create(save_dir, recursive = TRUE)
#setwd("/Users/hungm/The Francis Crick Dropbox/CaladoD/Matthew/github-repos/r-pipelines/deseq2pip/tests/rnaseq/module/")
```

## Data Import and Setup

Let's import and prepare your data. We'll use the example dataset GSE189410 to demonstrate the workflow:
```{r}
#| vscode: {languageId: r}
# load example DESeq2 object
rdata <- system.file("data", "GSE189410.dds.RData", package = "deseq2pip")
tx2gene <- gzfile(system.file("data", "GSE189410.tx2gene.tsv.gz", package = "deseq2pip"))
dds <- import_nfcore_rna(rdata = rdata, tx2gene = tx2gene)


# set group variable (optional to manually set reference level)
dds$Group2 <- factor(dds$Group2, c('IgM', 'IgG', 'IgA'))

# set dds design with desired variables
# for more information about setting contrasts in DESeq2, see: https://github.com/tavareshugo/tutorial_DESeq2_contrasts?tab=readme-ov-file
design(dds) <- ~ Group2
```


## Detailed Overview of DESeq2 pipeline

The run_deseq2_pip() function is a wrapper function that runs the complete RNA-seq analysis for **every PAIRWISE and ONE-TO-ALL comparison pairs** in the specified group variable.

The analysis for each comparison pair can be broken down into three modules:

1. Quality Control: Filters genes, generates QC/PCA/distance plots, and saves expression data
* `run_qc_pip()`
* `run_dist_pip()`
2. Differential Expression Analysis: Performs DESeq2 analysis, functional gene annotation, and creates volcano plots
* `run_diffexp_pip()`
* `plot_peak_annot_pip()` (ATAC-seq only)
3. GSEA Analysis: Find enriched pathways using MSigDB collections and prepare results for EnrichmentMap visualization
* `run_gsea_pip()`
* `run_enrichmentmap_pip()`

For our example dataset, we have 3 comparison pairs in the group variable `Group2`: `IgM` vs `IgG`, `IgM` vs `IgA`, and `IgG` vs `IgA`.

Here's how to run the complete pipeline with a single command:
*Note: Repeating run_deseq2_pip will overwrite the existing files if the same directory name is used. To prevent this, please specify a different save_dir directory.*
```{r}
#| vscode: {languageId: r}
# run deseq2 pipeline
#run_rna_pip(dds, experiment = "GSE189410", org = "mouse", group_by = "Group2", remove_xy = TRUE, remove_mt = TRUE, save_dir = save_dir)
```

## Run individual modules

After an initial run of the DESeq2 pipeline, you may wish to run a specific module using different parameters or comparison groups without repeating the entire pipeline.

Here's how to run each module separately. First, set the global arguments.
```{r}
#| vscode: {languageId: r}
# set pipeline arguments
group_by <- "Group2"
org <- "mouse"
remove_xy <- TRUE
remove_mt <- TRUE
min_count <- 10
batch <- NULL
pals <- NULL
order <- "pxfc"
one_to_all <- FALSE
msigdbr <- import_msigdbr(org = org)

# set additional arguments
save_data <- FALSE
save_plot <- FALSE
```

Below are the commands to run each individual module:


```{r}
# Module 1: Quality Control
dds <- run_qc_pip(dds, org = org, group_by = group_by, remove_xy = remove_xy, remove_mt = remove_mt, min_count = min_count, save_dir = save_dir)
dds <- run_dist_pip(dds, group_by = group_by, batch = batch, pals = pals, save_dir = save_dir)
```

```{r}
# Module 2: Differential Expression Analysis
res <- run_diffexp_pip(dds, org = org, group_by = group_by, order = order, one_to_all = one_to_all, save_dir = save_dir)
#res <- plot_peak_annot_pip(res, org = org, save_dir = save_dir)
```

```{r}
# Module 3: GSEA Analysis
group_dir <- paste0(save_dir, "/pairwise_", group_by)
gsea <- run_gsea_pip(res = res, org = org, msigdbr = , group_save_dir = group_dir)
enrichmentmap_pip(dds, org = org, group_by = group_by, group_dir = group_dir, save_dir = group_dir)
```


## Module 1 Breakdown

The `run_qc_pip()` module contains a sequence of subprocesses that performs quality control checks on the data. This includes removing XY and MT genes, removing lowly expressed genes, and checking library size distribution.

### Remove XY and MT genes
If your data contains samples taken from both males and females that you are not interested in, you can optionally remove gender-specific genes on X and Y chromosomes. This can be done with the `remove_xy_genes()` function.

Similarly, if your data contains mitochondrial genes that you are not interested in, you can optionally remove them with the `remove_mt_genes()` function.

This is optional but recommended if you see a gender-specific effect in downstream PCA and distance analysis.

```{r}
#| vscode: {languageId: r}
# when remove_xy or remove_mt are TRUE
if(remove_xy) { 
  dds <- remove_xy_genes(dds, org = org)}
if(remove_mt) {
  dds <- remove_mt_genes(dds, org = org)}
```


### Remove lowly expressed genes
RNA-seq is a high-throughput technology that can be noisy, and certain genes are poorly captured from RNA-seq. To control for this, we will remove these genes to improve statistical power for subsequent analysis.

By default we keep genes with at least 10 raw counts (`remove_low_expression(..., min_count = 10)`). The gene must also be expressed in at least n replicates, where n is the minimum replicates in the group.

```{r}
#| vscode: {languageId: r}
# remove low expression genes
dds <- remove_low_expression(dds, min_count = min_count, group_by = group_by, save_plot = save_plot, save_dir = save_dir)
```


### Check library size distribution
Expression of genes may vary between samples, and this may be due to sample/batch quality issues. 

Here we draw a boxplot representing expression level of every gene and compare the gene expression distribution from each sample using the `check_library()` function.

If the boxplots show little variation in expression level across samples, this indicates that the samples are of high quality and suitable for differential expression analysis. 

Otherwise, one should consider removing the samples with low quality or perform batch correction downstream to address quality issues. This will address technical variation between samples/batches and improve the statistical power for true biological variations.

```{r}
#| vscode: {languageId: r}
# check library size distribution
check_library(dds, save_plot = save_plot, save_dir = save_dir)
```


### Run PCA and distance analysis 
Next we visualize the relationship between samples/groups using PCA and distance analysis using the `run_pca()` and `run_distance()` functions. 

If there are true biological variability between groups, replicas from the same group should cluster closely in the PCA plot and have lower distance values in the distance heatmap.

Again, one should consider sample/batch quality if the samples do not separate as expected.

```{r}
#| vscode: {languageId: r}
# run PCA for the samples with group information
vsd <- vst(dds, blind = FALSE)
run_pca(vsd, group_by = group_by, size = 4, save_data = save_data, save_plot = save_plot, save_dir = save_dir)

# run distance analysis between samples
run_distance(vsd, save_data = save_data, save_plot = save_plot, save_dir = save_dir)
```


## Module 2: Differential Expression Analysis

The `run_diffexp_pip()` module performs statistical analysis to identify differentially expressed genes between groups using `DESeq2::DESeq2()` function, with p-values were generated using DESeq2 test. 

### Main function

The main function in the module is the `run_diffexp_pip()` function, which by default will perform all pairwise/one-to-all comparisons between groups in the DESeq2 object.

This returns a single dataframe all metrics from `DESeq2::results()`, and the differentially expressed genes for each comparison is separated by `comparison` column. 

Fold change shrinkage is also performed by `ashr` method. 

Additionally, chromosome loci and functional annotations of each gene was added to the dataframe using the `strpip::run_annotation()` function (see https://github.com/hungms/strpip).

```{r}
#| vscode: {languageId: r}
# run differential expression analysis for all comparisons; skipped here
#res <- run_diffexp_pip(dds, group_by = group_by, one_to_all = one_to_all, save_data = save_data, save_dir = save_dir)
#head(res)

# run differential expression analysis for a specific comparison
res <- run_diffexp(dds, org = org, group_by = group_by, comparison = "IgG_vs_IgA", order = order, save_data = save_data, save_dir = save_dir)
head(res)
```

### Generating MA & volcano plots

Here we will use `plot_ma()` and `plot_volcano()` to generate a single MA plot for the `IgG_vs_IgA` comparison as a demonstration.
```{r}
# generate MA plot for the "IgG_vs_IgA" comparison
plot_ma(res, save_plot = save_plot, save_dir = save_dir)

# Generate volcano plot for the "IgA_vs_IgG" comparison
plot_volcano(res, save_plot = FALSE, save_dir = save_dir)
```


## Module 3: GSEA Analysis

The final `run_gsea_pip()` module performs gene set enrichment analysis to identify biological pathways and processes using a list of MSigDB gene set collections (HALLMARK, GOBP, KEGG, REACTOME, TFT, BIOCARTA). Gene sets must have at least 10 and no more than 1000 genes to be tested.


### Main function

This returns a list of gseaResult class objects for each gene set collection and comparison.

```{r}
#| vscode: {languageId: r}
# run gsea analysis
# gsea <- run_gsea_pip(res, org = org, save_data = save_data, save_dir = save_dir)

# run gsea analysis
gsea <- run_gsea(res, org = org, save_data = save_data, save_dir = save_dir)
names(gsea)
```

### Generate barplots for top enriched gene set terms
Here we will use `plot_gsea_barplot()` to generate a single barplot for the `IgA_vs_IgG` comparison using GOBP collection as a demonstration. By default, the barplot will show the top 10 upregulated and downregulated terms.
```{r}
# generate barplot for the "IgA_vs_IgG" comparison
plot_gsea_barplot(gsea, n = 10, save_plot = FALSE, save_dir = save_dir)
```


### EnrichmentMap pipeline

Finally the module will run `enrichmentmap_pip()` which creates a enrichmentmap directory containing files formated for EnrichmentMap visualization.

By default, the function will only use the HALLMARK, GOBP, KEGG and REACTOME collections. We will skip this step here.

```{r}
# run enrichmentmap pipeline
#enrichmentmap_pip(gsea, org = org, group_by = group_by, group_dir = group_dir, save_dir = group_dir)
```